Discrete and continuous latent states of neural activity in \textit{Caenorhabditis elegans}

### 2017-11-09 drafts/2017_11_06/zimmer.tex
final draft of what was sent to zimmer

\begin{abstract}
  Recent advances in neural recording technologies have enabled
  simultaneous measurements of the majority of head ganglia neurons in
  both immobilized and freely-behaving \celegans.  The dynamics of
  neural activity shed light on how \celegans~processes sensory
  information and generates motor activity.  To understand these
  dynamics, we leverage \emph{recurrent switching linear dynamical
    systems}~\citep{linderman2017recurrent}---probabilistic models
  that decompose complex time-series into segments with simple, linear
  dynamics. We incorporate these models into a robust, hierarchical
  framework for combining information across whole-brain recordings of
  multiple worms.  We use this framework to reveal latent states of
  population neural activity, along with the discrete modes of
  behavior that drive dynamics in this latent state space.  We find
  characeristic transition patterns between these discrete modes, and
  we see that transition probabilities are determined, in part, by the
  current brain state.  This probabilistic framework is a useful aid
  for neural identification, currently a laborious, manual task.
  Moreover, we find clusters of neurons that are similarly tuned in
  latent state space, and we show how the different dynamical modes
  activate these clusters.  Finally, we find a significant overlap
  between our inferred modes and the manually-labeled states
  of~\citet{kato2015global}, which were shown to correspond to
  different worm behaviors. Our framework automatically discovers
  these behaviorally-meaningful regimes directly from neural activity.
\end{abstract}

@@

The data comes in the form of a collection of matrices~$\{\widetilde{Y}^{(w)}\}_{w=1}^W$
for each of~$W$ worms.  The matrix~${\widetilde{Y}^{(w)} \in \reals^{T_w \times N_w}}$
representes the calcium activity of the~$w$-th worm, which consists of~$T_w$
time frames and~$N_w$ neurons.\footnote{Technically, this calcium activity is
a matrix of first-order temporal differences of a
bleaching-corrected~$\Delta F /F$ signal. Unlike~\citet{kato2015global}, we
do not smooth these derivates. We work with the raw first-order
temporal differences.}  \celegans~is special in that each neuron has a unique
name.  Assigning names to neurons in calcium imaging data is a challenging
task, but typically, out of the~${\sim 100}$ neurons observed in any worm,
we can label about~$20-30$ with high certainty.  As such, we choose to
represent each matrix~$\widetilde{Y}^{(w)}$ in \emph{canonical form} as a
tuple~$(Y^{(w)}, M^{(w)})$, where~${Y^{(w)} \in \reals^{T_w \times N}}$ is
a matrix where each column corresponds to one of the~$N$ neurons that is
identified in at least one of the~$W$ worms, and~${M^{(w)} \in \{0,1\}^{T_w \times N}}$
is a corresponding \emph{mask} matrix that indicates which neurons were
observed in worm~$w$.  If~$M_{:,n}^{(w)} = 1$, neuron~$n$ was labeled in
worm~$w$ and~$Y_{:,n}^{(w)}$ was its observed activity.  If~$M_{:,n}^{(w)}=0$,
this neuron was not observed in worm~$w$, and the corresponding activity
is missing.\footnote{The mask is defined in such a way that neurons can
  be missing for only subsets of time frames, but in our data neurons are
  either labeled or missing.}
% todo: note that we could also (and have also) model the activity of
% unlabeled neurons.  For the final analyses, I'm not doing this though.

@@

\paragraph{Recurrent model of discrete state transition probabilities.}
Finally, we consider a third extension that allows for more complex
transition probabilities.  We replace~\eqref{eq:markov_z} with,
\begin{align}
  z_{t+1}^{(w)} \given z_t^{(w)}, x_t^{(w)}
  &\sim \mathrm{softmax}(r_{z_t^{(w)}} + R x_t^{(w)}),
    \label{eq:recurrent}
\end{align}
where~${r_k \in \reals^K}$ and~${R \in \reals^{K \times D}}$
parameterize a map from previous discrete and continuous states to a
distribution over next discrete states.  The link function is the
softmax function, which exponentiates and noramlizes
the~$K$-dimensional argument~${r_{z_t^{(w)}} + R x_t^{(w)}}$.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=6in]{figures/lds/dimensionality} \\
  \includegraphics[width=5.5in]{figures/lds/observation_variance} 
  \caption{\textit{(top)} Training and test likelihoods for the linear
    dynamical system and its hierarchical extension as a function of
    the dimensionality of the latent state,~$D$ (units:
    nats/observation over Gaussian baseline). On the basis of held-out
    test likelihood, we choose the hierarchical LDS with
    10-dimensional latent states.  \textit{(bottom)} Inferred standard
    deviaion~$\sqrt{s_n^{(w)}}$ for each neuron and worm.  Gray cells
    indicate neurons that were not observed in that worm.}
  \label{fig:lds}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=5.5in]{figures/lds/y_0}
  \caption{The observed activity~$y$ (i.e. the first-order differences
    of the calcium signal) are shown in black; the smoothed activity
    with the hierarchical LDS is shown in blue.  The hierarchical LDS
    makes predictions about unobserved neurons---those with dotted
    black lines---based on the correlations found in other worms. We
    show only one minute of data and the first 20 neurons
    (alphabetically) here.}
  \label{fig:activity_1}
\end{figure}

Figures~\ref{fig:activity_1}-\ref{fig:activity_5} show the observed
activity~$y$ along with the smoothed activity under the hierarchical
LDS.  Formally, the blue lines are given by~$C \E[x \given y] +
d$. Unobserved neurons are shown as dotted black lines.  Though their
activity is not observed, the hierarchical model can predict it based
on correlations observed in other neurons. We note that some of the
highly active neurons (e.g. \textsf{AVFL} and \textsf{AVFR}) are
largely smoothed out, as the model attributes much of their activity
to noise.

@@

The rows of the emission matrix~$c_n$ serve as ``embeddings'' of the
neurons. These embeddings can aid in this matching.  Specifically,
given an activity trace~$y^{(w)}_{n^*} \in \reals^T$ for an unlabled
neuron~$n^*$ in worm~$w$, along with the set of continuous latent
states~$x^{(w)} \in \reals^{T \times D}$ inferred from the labeled
neurons, we perform a linear regression to
estimate~$c_{n^*} \in \reals^D$, the ``embedding'' of this unlabeled
neuron.  We then compare this vector~$c_{n^*}$ to the embeddings of
the labeled neurons,~$\{c_n\}_{n=1}^N$, which were learned with the
SLDS.  We use the cosine similarity between two vectors as a measure
of likeness.

\begin{figure}[h]
\centering%
\includegraphics[width=3in]{figures/lds/similarity_comparison} 
\caption{The emission parameters aid in neural identification.  In
  this task, we view unlabeled neurons as \emph{candidates} for a
  particular label. For example, suppose we are trying to determine
  the candidate neuron that is most likely to be~\textsf{AVAL}. For
  this task, we have artifically held-out this label so that we know
  which candidates truly is~\textsf{AVAL}. We use the neural
  embeddings~$c_{\mathsf{AVAL}}$, which was inferred from other worms,
  to aid in this matching.  We measure the cosine similarity
  between~$c_{\mathsf{AVAL}}$ and the coefficients of linear
  regressions between~$x$ and the activity of the unlabeled neurons'
  activity. This histogram shows the results of performing this
  experiment with 10 held-out identities for each of the five worms.
  We see that the correct candidates frequently have much higher
  cosine similarity than incorrect candidates.  This means that we can
  use this similarity measure to infer neuron identities.  }
\label{fig:neuron_id}
\end{figure}

@@

Figure~\ref{fig:rec_weights} shows the two terms that go into the
distribution over next states: the ``Markov'' term~$\{r_{k}\}_{k=1}^K$
and the ``recurrent'' weights~$R$, as defined in
Eq.~\eqref{eq:recurrent}. The largest weights come from the
self-transitions, i.e. the diagonal of the Markov terms. Still, the
recurrent weights are non-negligible. The histogram of Markov
weights~($r_{z_t}$) and the recurrent weights~($R^\trans x_t$) at the
times of transitions (right panel) shows that the recurrent weights
are comparable to the Markov weights, and therefore must play a role
in determining transition probabilities.  This must be the case as the
recurrent models yield significant improvement in held-out log
likelihoods, as shown in Figure~\ref{fig:arhmm_lls}.

@@

\subsection{Some state durations are non-Markovian}

\begin{figure}[h]
\centering%
\includegraphics[width=1.8in]{figures/arhmm/durations_0.pdf}
\includegraphics[width=1.8in]{figures/arhmm/durations_1.pdf}
\includegraphics[width=1.8in]{figures/arhmm/durations_2.pdf}
\\
\includegraphics[width=1.8in]{figures/arhmm/durations_3.pdf}
\includegraphics[width=1.8in]{figures/arhmm/durations_4.pdf}
\includegraphics[width=1.8in]{figures/arhmm/durations_5.pdf}
\\
\includegraphics[width=1.8in]{figures/arhmm/durations_6.pdf}
\includegraphics[width=1.8in]{figures/arhmm/durations_7.pdf}
\caption{Empirical distribution of state durations (histogram)
  compared to the geometric distribution implied by the Markovian
  weights of the transition model.  Some state durations are
  well-approximated by the geometric distribution, but others
  are peaked away from zero (e.g. states 2, 4, 6, and 7).  The
  recurrent transition model allows the AR-HMM to capture this
  non-Markovianity by modeling transitions as a function of location
  in latent state space~$x_t$.}
\label{fig:durations}
\end{figure}

Figure~\ref{fig:durations} shows histograms of the distribution of inferred
state distributions for each state.  For comparison, we show
the geometric distribution of state durations implied by the
Markov weights as a black line.  Specifically, the distribution
over durations~$d_k$ of state~$k$ under the Markov model (ignoring the
recurrent weights) is given by,
\begin{align*}
  p_{k \to k} &= [\mathrm{softmax}(r_{k})]_k, \\
  d_k &\sim \mathrm{Geom}(1-p_{k \to k}).
\end{align*}
We see that some states (e.g. states 1, 5, and 8) seem to follow the
Markov assumption quite well, whereas others (states 2, 4, and especially 7),
are far from Markovian.  The recurrent model incorporates the continuous
latent state~$x_t$ into the transition probabilities via the additive term~$R x_t$
in the softmax.  It is likely that this term captures the non-Markovianity.

@@

A few patterns emerge here:
\begin{itemize}
\item Worm 3 differs from the rest, and seems to have a separate
  matching of inferred states and true states.  This list focuses
  on patterns in Worms 1, 2, 4, and 5. 
\item \textsf{REVSUS} --- the most common manually labeled state ---
  tends to overlap with the red and yellow inferred states.
\item \textsf{SLOW} --- the second most common --- and \textsf{FWD}
  often map onto the blue state.
\item \textsf{VT} (ventral turns) map to the green state, which
  has a rise and fall of activity in cluster 3 (neurons
  {\sf ALA, ASKL, ASKR, RIVL, RIVR, SMDVL} and \textsf{SMDVR}).
  This seems consistent with the roles of sensory neurons~\textsf{ASK*},
  interneurons \textsf{RIV*}, and motor neurons \textsf{SMDV*} in
  local search behavior.
\item Skipping ahead, \textsf{DT} (dorsal turns) often map onto
  the orange and, to a much lesser extent, the purple state. Both
  states encourage upswings in cluster 4 ({\sf SMBDL, SMBDR, SMBVL, SMBVR}, and \textsf{URYVR}).
  The first four of these neurons are motor neurons involved in
  local search behavior. 
\item In addition to the blue state, \textsf{FWD} maps onto the
  green, purple, and orange states as well, though much less significantly.
\item \textsf{REV2} and \textsf{REV1} map primarily onto the pink state.
  This state is characterized by an excursion of activity along cluster
  1 ({\sf AVAL, AVAR, BAGL, RIML, RIMR, SAAVR, SABVL, SABVR, VA01}.
  The \textsf{AVA*} interneurons innvervate motor neurons.
  \textsf{RIM*} are motor neurons that modulate reversals.
  \textsf{SABV*} are interneurons and sublateral motor neurons.
  \textsf{VA01} is a motor neuron.

@@

\subsection{Inferred changepoints also match manual labels}

\begin{figure}[h]
\centering%
\includegraphics[width=5.5in]{figures/arhmm/z_cps_worm0.pdf}
\includegraphics[width=5.5in]{figures/arhmm/z_cps_worm1.pdf}
\caption{(top) Posterior distribution over latent states~$z_t$.
  Each column corresponds to a time bin and sums to one.
  (middle) The posterior probability of changepoints.
  (bottom) For reference,
  the manually labeled states of~\citet{kato2015global}. Colors match those
  of Figure~\ref{fig:overlap}.}
\label{fig:z_cps_12}
\end{figure}

Figures~\ref{fig:z_cps_12}, \ref{fig:z_cps_34}, and \ref{fig:z_cps_5}
show the posterior distribution of latent states~$z$ inferred from
neural activity alone. From this posterior, we also compute the probability
of a changepoint, where~$z_{t+1} \neq z_t$.  For reference, we show the
manually labeled states of~\citet{kato2015global} below. Here, patterns
like the cascade from states 2 to 3 to 4 (red to yellow to green) are
apparent.

\begin{figure}[h]
\centering%
\includegraphics[width=5.5in]{figures/arhmm/z_cps_worm2.pdf}
\includegraphics[width=5.5in]{figures/arhmm/z_cps_worm3.pdf}
\caption{See Figure~\ref{fig:z_cps_12} caption.}
\label{fig:z_cps_34}
\end{figure}

\begin{figure}[h]
\centering
\includegraphics[width=5.5in]{figures/arhmm/z_cps_worm4.pdf}
\caption{See Figure~\ref{fig:z_cps_12} caption.}
\label{fig:z_cps_5}
\end{figure}

\subsection{Inferred states changepoints correspond with changes in dynamics}

Figures~\ref{fig:segmentation_1} through~\ref{fig:segmentation_5} show an
interval of latent state trajectories~$x$ along with the segmentation from
the inferred discrete states~$z$ (top).  For comparison, we also show the
manual segmentation of~\citep{kato2015global} (bottom).  The inferred
segmentation seems to identify changepoints when one or more of the latent
continuous states changes dynamics.  Indeed, this is the assumption of the
model.  By contrast, the segmentation of~\citep{kato2015global} is based on
a read-out of a single neuron, and while it is largely consistent with the
inferred segmentation at the population level, it misses some salient
changes in dynamics.  For example, see the inferred changepoint around
time 700 in worm 1.  This changepoint was not identified in the manual
data.

>>>>

\begin{abstract}
  Recent advances in neural recording technologies have enabled
  simultaneous measurements of the majority of head ganglia neurons in
  both immobilized and freely-behaving \celegans.  The dynamics of
  neural activity shed light on how \celegans~processes sensory
  information and generates motor activity.  To understand these
  dynamics, we leverage \emph{recurrent switching linear dynamical
    systems}~\citep{linderman2017recurrent}---probabilistic models
  that decompose complex time-series into segments with simple, linear
  dynamics. We incorporate these models into a robust, hierarchical
  framework for combining information across whole-brain recordings of
  multiple worms.  We use this framework to reveal latent states of
  population neural activity, along with the discrete modes of
  behavior that drive dynamics in this latent state space.  We find
  characteristic transition patterns between these discrete modes, and
  we see that transition probabilities are determined, in part, by the
  current brain state.  This probabilistic framework is a useful aid
  for neural identification, currently a laborious, manual task.
  Moreover, we find clusters of neurons that are similarly tuned in
  latent state space, and we show how the different dynamical modes
  activate these clusters.  Finally, we find a significant overlap
  between our inferred modes and the manually-labeled states
  of~\citet{kato2015global}, which were shown to correspond to
  different worm behaviors. Our framework automatically discovers
  these behaviorally-meaningful regimes directly from neural activity.
\end{abstract}

@@

The data comes in the form of a collection of matrices~$\{\widetilde{Y}^{(w)}\}_{w=1}^W$
for each of~$W$ worms.  The matrix~${\widetilde{Y}^{(w)} \in \reals^{T_w \times N_w}}$
represents the calcium activity of the~$w$-th worm, which consists of~$T_w$
time frames and~$N_w$ neurons.\footnote{Technically, this calcium activity is
a matrix of first-order temporal differences of a
bleaching-corrected~$\Delta F /F$ signal. Unlike~\citet{kato2015global}, we
do not smooth these derivatives. We work with the raw first-order
temporal differences.}  \celegans~is special in that each neuron has a unique
name.  Assigning names to neurons in calcium imaging data is a challenging
task, but typically, out of the~${\sim 100}$ neurons observed in any worm,
we can label about~$20-30$ with high certainty.  As such, we choose to
represent each matrix~$\widetilde{Y}^{(w)}$ in \emph{canonical form} as a
tuple~$(Y^{(w)}, M^{(w)})$, where~${Y^{(w)} \in \reals^{T_w \times N}}$ is
a matrix where each column corresponds to one of the~$N$ neurons that is
identified in at least one of the~$W$ worms, and~${M^{(w)} \in \{0,1\}^{T_w \times N}}$
is a corresponding \emph{mask} matrix that indicates which neurons were
observed in worm~$w$.  If~$M_{:,n}^{(w)} = 1$, neuron~$n$ was labeled in
worm~$w$ and~$Y_{:,n}^{(w)}$ was its observed activity.  If~$M_{:,n}^{(w)}=0$,
this neuron was not observed in worm~$w$, and the corresponding activity
is missing.\footnote{The mask is defined in such a way that neurons can
  be missing for only subsets of time frames, but in our data neurons are
  either labeled or missing.}
% todo: note that we could also (and have also) model the activity of
% unlabeled neurons.  For the final analyses, I'm not doing this though.

@@

\paragraph{Recurrent model of discrete state transition probabilities.}
Finally, we consider a third extension that allows for more complex
transition probabilities.  We replace~\eqref{eq:markov_z} with,
\begin{align}
  z_{t+1}^{(w)} \given z_t^{(w)}, x_t^{(w)}
  &\sim \mathrm{softmax}(r_{z_t^{(w)}} + R x_t^{(w)}),
    \label{eq:recurrent}
\end{align}
where~${r_k \in \reals^K}$ and~${R \in \reals^{K \times D}}$
parameterize a map from previous discrete and continuous states to a
distribution over next discrete states.  The link function is the
softmax function, which exponentiates and normalizes
the~$K$-dimensional argument~${r_{z_t^{(w)}} + R x_t^{(w)}}$.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=6in]{figures/lds/dimensionality} \\
  \includegraphics[width=5.5in]{figures/lds/observation_variance} 
  \caption{\textit{(top)} Training and test likelihoods for the linear
    dynamical system and its hierarchical extension as a function of
    the dimensionality of the latent state,~$D$ (units:
    nats/observation over Gaussian baseline). On the basis of held-out
    test likelihood, we choose the hierarchical LDS with
    10-dimensional latent states.  \textit{(bottom)} Inferred standard
    deviation~$\sqrt{s_n^{(w)}}$ for each neuron and worm.  Gray cells
    indicate neurons that were not observed in that worm.}
  \label{fig:lds}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=5.5in]{figures/lds/y_0}
  \caption{The observed activity~$y$ (i.e. the first-order differences
    of the calcium signal) are shown in black; the smoothed activity
    with the hierarchical LDS is shown in blue.  The hierarchical LDS
    makes predictions about unobserved neurons---those with dotted
    black lines---based on the correlations found in other worms. We
    show two minutes of data here.  The neurons have been grouped into
    five clusters, as discussed below.  Within each cluster, the
    neurons are sorted alphabetically. Each neuron is individually
    scaled by the standard deviation of its observed activity to
    illustrate dynamics of low-amplitude neurons. }
  \label{fig:activity_1}
\end{figure}

Figures~\ref{fig:activity_1}-\ref{fig:activity_5} show the observed
activity~$y$ along with the smoothed activity under the hierarchical
LDS.  Formally, the blue lines are given by~$C \E[x \given y] +
d$. Unobserved neurons are shown as dotted black lines.  Though their
activity is not observed, the hierarchical model can predict it based
on correlations observed in other worms.

@@

The rows of the emission matrix~$c_n$ serve as ``embeddings'' of the
neurons. These embeddings can aid in this matching.  Specifically,
given an activity trace~$y^{(w)}_{n^*} \in \reals^T$ for an unlabeled
neuron~$n^*$ in worm~$w$, along with the set of continuous latent
states~$x^{(w)} \in \reals^{T \times D}$ inferred from the labeled
neurons, we perform a linear regression to
estimate~$c_{n^*} \in \reals^D$, the ``embedding'' of this unlabeled
neuron.  We then compare this vector~$c_{n^*}$ to the embeddings of
the labeled neurons,~$\{c_n\}_{n=1}^N$, which were learned with the
SLDS.  We use the cosine similarity between two vectors as a measure
of likeness.

\begin{figure}[h]
\centering%
\includegraphics[width=3in]{figures/lds/similarity_comparison} 
\caption{The emission parameters aid in neural identification.  In
  this task, we view unlabeled neurons as \emph{candidates} for a
  particular label. For example, suppose we are trying to determine
  the candidate neuron that is most likely to be~\textsf{AVAL}. For
  this task, we have artificially held-out this label so that we know
  which candidates truly is~\textsf{AVAL}. We use the neural
  embeddings~$c_{\mathsf{AVAL}}$, which was inferred from other worms,
  to aid in this matching.  We measure the cosine similarity
  between~$c_{\mathsf{AVAL}}$ and the coefficients of linear
  regressions between~$x$ and the activity of the unlabeled neurons'
  activity. This histogram shows the results of performing this
  experiment with 10 held-out identities for each of the five worms.
  We see that the correct candidates frequently have much higher
  cosine similarity than incorrect candidates.  This means that we can
  use this similarity measure to infer neuron identities.  }
\label{fig:neuron_id}
\end{figure}

@@

Figure~\ref{fig:rec_weights} shows the two terms that go into the
distribution over next states: the ``Markov'' term~$\{r_{k}\}_{k=1}^K$
and the ``recurrent'' weights~$R$, as defined in
Eq.~\eqref{eq:recurrent}.  Since the impact of the recurrent weights
depends on the scale of the continuous latent states, we multiply the
weight matrix~$R$ by the standard deviation of~$x$ (since~$x$ is
roughly zero mean, this is like z-scoring~$x$ so that it is comparable
with the Markov term).  The largest weights come from the
self-transitions, i.e. the diagonal of the Markov terms. Still, the
recurrent weights are non-negligible. The histogram of Markov
weights~($r_{z_t}$) and the recurrent weights~($R^\trans x_t$) at the
times of transitions (right panel) shows that the recurrent weights
are comparable to the Markov weights, and therefore must play a role
in determining transition probabilities.  This must be the case as the
recurrent models yield significant improvement in held-out log
likelihoods, as shown in Figure~\ref{fig:arhmm_lls}.

@@

\subsection{The recurrent model captures non-Markovian state durations}

\begin{figure}[h]
\centering%
\includegraphics[width=1.8in]{figures/arhmm/durations_0.pdf}
\includegraphics[width=1.8in]{figures/arhmm/durations_1.pdf}
\includegraphics[width=1.8in]{figures/arhmm/durations_2.pdf}
\\
\includegraphics[width=1.8in]{figures/arhmm/durations_3.pdf}
\includegraphics[width=1.8in]{figures/arhmm/durations_4.pdf}
\includegraphics[width=1.8in]{figures/arhmm/durations_5.pdf}
\\
\includegraphics[width=1.8in]{figures/arhmm/durations_6.pdf}
\includegraphics[width=1.8in]{figures/arhmm/durations_7.pdf}
\caption{Empirical distribution of state durations (histogram)
  compared to the geometric distribution implied by the Markovian
  weights of the transition model.  Some state durations are
  well-approximated by the geometric distribution, but others
  are peaked away from zero (e.g. states 2, 4, 6, and 7).  The
  recurrent transition model allows the AR-HMM to capture this
  non-Markovianity by modeling transitions as a function of location
  in latent state space~$x_t$. Solid lines show the distribution
  of simulated durations under the recurrent AR-HMM; dashed lines
  show the geometric duration distribution under the Markovian model.}
\label{fig:durations}
\end{figure}

Figure~\ref{fig:durations} shows histograms of the
empirical distribution of inferred state durations for each state. For
comparison, the solid lines show the distribution of
\emph{simulated} state distributions from our recurrent AR-HMM.  We
see a close correspondence.

This correspondence is a direct result of the recurrent nature of the
model.  Without the recurrent terms, the state duration distributions
would necessarily follow the geometric distribution shown in dotted
black lines.  These distributions decay monotonically and cannot
capture the peaks away from zero that we find in the data.  This is
due to the fact that the distribution over durations~$d_k$ of
state~$k$ under a Markov model (ignoring the recurrent weights) is
given by,
\begin{align*}
  p_{k \to k} &= [\mathrm{softmax}(r_{k})]_k, \\
  d_k &\sim \mathrm{Geom}(1-p_{k \to k}).
\end{align*}
While some states (e.g. states 1, 5, and 8) seem to follow the Markov
assumption quite well, others (states 2, 4, and especially 7), are far
from Markovian.  The recurrent model incorporates the continuous
latent state~$x_t$ into the transition probabilities via the additive
term~$R x_t$ in the softmax and thereby captures the non-Markovianity.

@@

A few patterns emerge here:
\begin{itemize}
\item Worm 3 differs from the rest, and seems to have a separate
  matching of inferred states and true states.  This list focuses
  on patterns in Worms 1, 2, 4, and 5. 
\item \textsf{REVSUS} --- the most common manually labeled state ---
  tends to overlap with the red and yellow inferred states.
\item \textsf{SLOW} --- the second most common --- and \textsf{FWD}
  often map onto the blue state.
\item \textsf{VT} (ventral turns) map to the green state, which
  has a rise and fall of activity in cluster 3 (neurons
  {\sf ALA, ASKL, ASKR, RIVL, RIVR, SMDVL} and \textsf{SMDVR}).
  This seems consistent with the roles of sensory neurons~\textsf{ASK*},
  interneurons \textsf{RIV*}, and motor neurons \textsf{SMDV*} in
  local search behavior.
\item Skipping ahead, \textsf{DT} (dorsal turns) often map onto
  the orange and, to a much lesser extent, the purple state. Both
  states encourage upswings in cluster 4 ({\sf SMBDL, SMBDR, SMBVL, SMBVR}, and \textsf{URYVR}).
  The first four of these neurons are motor neurons involved in
  local search behavior. 
\item In addition to the blue state, \textsf{FWD} maps onto the
  green, purple, and orange states as well, though much less significantly.
\item \textsf{REV2} and \textsf{REV1} map primarily onto the pink state.
  This state is characterized by an excursion of activity along cluster
  1 ({\sf AVAL, AVAR, BAGL, RIML, RIMR, SAAVR, SABVL, SABVR, VA01}.
  The \textsf{AVA*} interneurons innervate motor neurons.
  \textsf{RIM*} are motor neurons that modulate reversals.
  \textsf{SABV*} are interneurons and sublateral motor neurons.
  \textsf{VA01} is a motor neuron.

@@

% \subsection{Inferred change-points also match manual labels}

% \begin{figure}[h]
% \centering%
% \includegraphics[width=5.5in]{figures/arhmm/z_cps_worm0.pdf}
% \includegraphics[width=5.5in]{figures/arhmm/z_cps_worm1.pdf}
% \caption{(top) Posterior distribution over latent states~$z_t$.
%   Each column corresponds to a time bin and sums to one.
%   (middle) The posterior probability of change-points.
%   (bottom) For reference,
%   the manually labeled states of~\citet{kato2015global}. Colors match those
%   of Figure~\ref{fig:overlap}.}
% \label{fig:z_cps_12}
% \end{figure}

% Figures~\ref{fig:z_cps_12}, \ref{fig:z_cps_34}, and \ref{fig:z_cps_5}
% show the posterior distribution of latent states~$z$ inferred from
% neural activity alone. From this posterior, we also compute the probability
% of a change-point, where~$z_{t+1} \neq z_t$.  For reference, we show the
% manually labeled states of~\citet{kato2015global} below. Here, patterns
% like the cascade from states 2 to 3 to 4 (red to yellow to green) are
% apparent.

% \begin{figure}[h]
% \centering%
% \includegraphics[width=5.5in]{figures/arhmm/z_cps_worm2.pdf}
% \includegraphics[width=5.5in]{figures/arhmm/z_cps_worm3.pdf}
% \caption{See Figure~\ref{fig:z_cps_12} caption.}
% \label{fig:z_cps_34}
% \end{figure}

% \begin{figure}[h]
% \centering
% \includegraphics[width=5.5in]{figures/arhmm/z_cps_worm4.pdf}
% \caption{See Figure~\ref{fig:z_cps_12} caption.}
% \label{fig:z_cps_5}
% \end{figure}

% \clearpage

\subsection{Inferred states change-points correspond with changes in dynamics}

Figures~\ref{fig:segmentation_1} through~\ref{fig:segmentation_5} show an
interval of latent state trajectories~$x$ along with the segmentation from
the inferred discrete states~$z$ (top).  For comparison, we also show the
manual segmentation of~\citep{kato2015global} (bottom).  The inferred
segmentation seems to identify change-points when one or more of the latent
continuous states changes dynamics.  Indeed, this is the assumption of the
model.  By contrast, the segmentation of~\citep{kato2015global} is based on
a read-out of a single neuron, and while it is largely consistent with the
inferred segmentation at the population level, it misses some salient
changes in dynamics.  For example, see the inferred change-point around
time 700 in worm 1.  This change-point was not identified in the manual
data.
